---
title: "Gradient Descent in R. Minimization of functions"
author: "Matt Bogard"
output:
  html_document: default
  html_notebook: default
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE, error=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, error = TRUE)
```

Source: http://econometricsense.blogspot.com/2011/11/gradient-descent-in-r.html

In a previous post I discussed the concept of gradient descent.  Given some recent work in the online machine learning course offered at Stanford,  I'm going to extend that discussion with an actual example using R-code  (the actual code is adapted from a computer science course at Colorado State, and the example is verbatim from the notes here:

http://www.cs.colostate.edu/~anderson/cs545/Lectures/week6day2/week6day2.pdf ) 

Suppose you want to minimize the function 1.2 * (x-2)^2 + 3.2. Basic calculus requires that we find the 1st derivative and solve for the value of x such that f'(x) = 0. This is easy enough to do, f'(x) = 2*1.2*(x-2). Its easy to see that a value of 2 satisfies f'(x) =  0. Given that the second order conditions hold, this is a minimum.

Its not alwasys the case that we would get a function so easy to work with, and in many cases we may need to numerically estimate the value that minimizes the function. Gradient descent offers a way to do this. Recall from my previous post the gradient descent algorithm can be summarized as follows:

repeat until convergence { \   
$\;\;\;\; X_{n+1} = X_n - \alpha \nabla F(X_n)$, or \   
$\;\;\;\; x:= x - \alpha \nabla F(x)$, or \   
$\;\;\;\; x \leftarrow x - \alpha \nabla F(x)$, or \   
} \   
(depending on your notational preferences)

Where $\nabla F(x)$  would be the derivative we calculated above for the function at hand and α is the learning rate. This can easily be implemented R. The following code finds the values of x that minimize the function above and plots the progress of the algorithm with each iteration. (as depicted in the image below)

```{r}
xs <- seq(0, 4,len=20) # create some values
 
# define the function we want to optimize
f <-  function(x) { 
  1.2 * (x-2)^2 + 3.2
}
 
# plot the function 
plot(xs, f (xs), 
     type = "l", 
     xlab = "x", 
     ylab = expression(1.2(x-2)^2 +3.2))
 
# calculate the gradient df/dx
grad <- function(x){
  1.2*2*(x-2)
}
 
 
# df/dx = 2.4(x-2), if x = 2 then 2.4(2-2) = 0
# The actual solution we will approximate with gradeint descent
# is  x = 2 as depicted in the plot below
 
lines (c(2,2), c(3,8), col="green", lty=2)         
text (2.1,7, "Closed-form solution", col="red", pos=4)



# gradient descent implementation
x <- 0.1       # initialize the first guess for x-value
xtrace <- x    # store x -values for graphing purposes (initial)
ftrace <- f(x) # store y-values (function eval at x) for graphing purposes (initial)

stepFactor <- 0.6 # learning rate 'alpha'

for (step in 1:100) {
  x <- x - stepFactor * grad(x) # gradient descent update
  xtrace <- c(xtrace, x) # update for graph
  ftrace <- c(ftrace, f(x)) # update for graph
}
 
lines (xtrace, ftrace, type="b", col="blue")
text (0.5, 6, "Gradient Descent", col="blue", pos= 4)
 
# print final value of x
print(x) # x converges to 2.0
```

```{r}
plot(xs, f (xs), xlim = c(1.5, 2.5))
lines (xtrace, ftrace, type="b", col="blue")
```
```{r}
plot(xs, f (xs), xlim = c(1.9, 2.1))
lines (xtrace, ftrace, type="b", col="blue")
```

