---
title: "Introductory Liner Algebra"
output:
  pdf_document: 
    latex_engine: xelatex
  html_notebook: default
---

```{r setup, include=FALSE, error=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, error = TRUE)
```

Source: https://www.dropbox.com/s/qcwknpvnvul4k7i/Introductory%20linear%20algebra%20with%20R%5BR%20article%20linear_algebra%5D.pdf?dl=0


## Addition of vectors
```{r}
a <- c(1, 3, 2)
b <- c(2, 8, 9)

a + b

```


## Inner product of vectors

$a · b = a_1b_1 + ... + a_nb_n$

```{r}
a * b
```


```{r}
sum(a * b)
```

> Note, that the inner product is a number – not a vector:

```{r}
crossprod(a, b)
```

## The length (norm) of a vector

$\| a\| = \sqrt{a.a}$

```{r}
sqrt(sum(a * a))
```

## The 0-vector and the 1-vector

```{r}
rep(0, 5)
```

```{r}
rep(1, 5)
```

## Orthogonal vectors

```{r}
v1 <- c(1, 1)
v2 <- c(-1, 1)

crossprod(v1, v2)
```

# Matrices

```{r}
# Create a matrix reading values by column (default)
A <- matrix(c(1, 3, 2, 2, 8, 9), ncol = 3)
A
```

```{r}
# create matrix reading valus by row
A2 <- matrix(c(1,3,2,2,8,9), ncol = 3,byrow = TRUE)
```

\  

Matrices can be created in different ways:

```{r}
# creating a matrix using concatenation of vectors
a <- c(1, 3, 2)
b <- c(2, 8, 9)

A3 <- matrix(c(a, b), ncol = 3, byrow = TRUE)
A3
```

```{r}
A3 <- rbind( c(1, 3, 2), 
             c(2, 8, 9)
)
A3
```

## Multiplying a matrix by a scalar

```{r}
7 * A
```

## Transpose of a matrix

```{r}
A            # original matrix

t(A)         # transpose
```

## Multiplication matrix by vectors

```{r}
A
```

```{r}
a
```

```{r}
A %*% a
```

But:
```{r}
A * a 
```
which is an element-wise multiplication.

## Multiplication of matrices

```{r}
A <- matrix(c(1,3,2,2,8,9), ncol=2)
B <- matrix(c(5,8,4,2), ncol=2)
A; B
A %*% B

```

# Special matrices

## Square matrix
```{r}
SQ <- matrix(1:9, nrow=3, ncol=3) 
SQ
```

```{r}
SQ2 <- matrix(1:16, nrow=4)
SQ2
```

```{r}
SQ3 <- matrix(rnorm(25), nrow=5)
SQ3
```

## Symmetric matrix

```{r}
S <- rbind(c(0, 1, 2),
           c(1, 0, 3),
           c(2, 3, 0)
           )
S
```
   
The transpose:
```{r}
tS <- t(S)
tS
```

Testing:
```{r}
S == tS
```
Using the `isSymmetric` function:
```{r}
isSymmetric(S)
```

## The 0-matrix

```{r}
matrix(0, nrow=3, ncol=3)
```

## The J matrix

```{r}
matrix(1, nrow=3, ncol=3)
```

## The diagonal matrix
```{r}
diag(c(1,2,3))
```

Applying `diag` to a matrix:
```{r}
SQ <- matrix(1:9, nrow=3, ncol=3) 
SQ
diag(SQ)
```

## The identity matrix
A diagonal matrix with 1s on the diagonal is called the identity matrix and is
denoted $I$ . The identity matrix satisﬁes that $I A = AI = A$. Likewise, if $x$ is a vector
then $I x = x$.

```{r}
diag(1, 3)
```

# The inverse matrix
```{r}
A <- matrix(c(1,3,2,4), ncol=2, byrow=T)
A
```

The inverse of $A$ is $A{-1}$ which can be obtained with R:
```{r}
Ai <- solve(A)
Ai
```

We prove with:

```{r}
A %*% Ai
```
which is the identity matrix.












# Making a symmetric matrix
```{r}
# create the matrix
sym <- matrix(rep(0, 3*3), nrow = 3)
sym

# show the upper and lower triangle
upper.tri(sym)
lower.tri(sym)
```

Fill the lower triangle:
```{r}
# fill the lower triangle
sym[lower.tri(sym)] <- seq(from=1, to=choose(3,1), by=1)
sym
```
Transpose the matrix:
```{r}
sym <- t(sym)
sym
```

Fill the upper triangle of the transposed matrix:
```{r}
sym[lower.tri(sym)] <- seq(from=1, to=choose(3,1), by=1)
sym


```

Check symmetry:
```{r}
# check symmetry
isSymmetric(sym)
```




## Building the `makeSymmetric` function
We use the function `choose` which is the combinatorial $_{n}\textrm{C}_r$ equal to:

$\frac{n!}{r! \, (n-r)!}$

which acounts for the lower or upper triangle matrix elements.


```{r}
makeSymmetric <- function(n) {
  s <- matrix(rep(0, n*n), nrow = n)                     # create matrix
  s[lower.tri(s)] <- seq(from=1, to=choose(n,2), by=1)   # fill the lower triangle
  s <- t(s)                                              # tranpose
  s[lower.tri(s)] <- seq(from=1, to=choose(n,2), by=1)   # fill lower triangle of transposed
  return(s)
}
```


```{r}
makeSymmetric(3)
makeSymmetric(4)
makeSymmetric(5)
isSymmetric(makeSymmetric(6))
```

