---
title: "Orthogonal matrices"
output:
  pdf_document: default
  html_notebook: default
---


```{r setup, include=FALSE, error=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, error = TRUE)
```



## Generating orthogonal matrices

Source: https://stat.ethz.ch/pipermail/r-help/2003-February/030388.html

\     
We create a 3x3 matrix:
```{r}
k <- 3
tstMat <- array(runif(k), dim=c(k,k))
tstMat
```

\   
Then, apply QR decomposition function `qr` to obtain the orthogonal matrix:
```{r}
tstOrth <- qr.Q(qr(tstMat))
tstOrth
```


Then, we proceed to prove that the resulting matrix is orthogonal by testing it with $O^T * O =I$:


```{r}
t(tstOrth) %*% tstOrth
```

Which is the identity matrix $I$.

## Example:
Here is another example from Wikipedia:

```{r}
a1 <- c(12, -51, 4)
a2 <- c(6, 167, -68)
a3 <- c(-4, 24, -41)

A <- matrix(c(a1, a2, a3), ncol = 3, byrow = T)
A
```

```{r}
q1 <- c(6/7, -69/175, -58/175)
q2 <- c(3/7, 158/175, 6/175)
q3 <- c(-2/7, 6/35, -33/35)

Q <- matrix(c(q1, q2, q3), ncol =3, byrow = T)
Q
```

```{r}
crossprod(A, Q)
```

```{r}
q1
q2
q3
```


```{r}
t(Q) %*% Q
```

Which is the identity matrix $I$.

## Generating a random orthogonal matrix with `pracma`
Source: https://artax.karlin.mff.cuni.cz/r-help/library/pracma/html/rortho.html


```{r include=F}
# install pracma package if required
if (!require("pacman")) install.packages("pacman")
pacman::p_load(pracma)
```



```{r}
# load library
library(pracma)

# Generate random orthogonal matrix
Q <- randortho(5)

```
\  

Test $Q Q^T$:

```{r}
zapsmall(Q %*% t(Q))    # rounding with zapsmall
```

\  

Test $Q^T Q$:
```{r}
zapsmall(t(Q) %*% Q)
```



   